

# Behavioral econometrics

## Example dataset

```{r}
library(foreign)
library(ggplot2)
library(dplyr)
library(tidyr)
Hey2001<- read.csv("../2021ECON5820/Hey2001.csv")

knitr::kable(head(Hey2001))
```

* decisions over 2 lotteries. $p$ or $q$
* $y=1$ iff $i$ chooses the q-lottery in decision $t$
* prizes: 0, 50, 100, 150 pounds

## Risk neutral

Maximization problem
$$
\max_l\{E[u(X\mid l)]\}
$$

Probabilistic choice function. Given choice set $\mathbb A$ (softmax)
$$
\Pr\left(l\right)\propto \exp\left(\lambda E[u(x)\mid l]\right)
$$



* As the utility of one action increases, I am more likely to take it.
* Interpretations
  + I make mistakes, but I recognize big payoff differences better than small ones
  + There is a random component to my utility (observed by the decision-maker but not econometrician)
  + There is a component of my utility that is not modeled well by $E[u(x)\mid l]$
  
Since we have only 2 choices, this becomes:
  
$$
\begin{aligned}
\Pr(y=1\mid p,q)&=\frac{\exp(\lambda E[u(x)\mid q])}{\exp(\lambda E[u(x)\mid q])+\exp(\lambda E[u(x)\mid p])}\\
&=\left(1+\exp\left(\lambda\left(E[u(x)\mid p]-E[u(x)\mid q]\right)\right)\right)^{-1}
\end{aligned}
$$

log-likeihood function:
$$
\mathcal L(\theta)=\sum_{t=1}^T\Pr(y_t\mid p_y,q_t)
$$

For risk-neutral: $u(x)=x$


```{r}

library(dplyr)

D<-(Hey2001
    %>% mutate(EVq = 0*q1+50*q2+100*q3+150*q4,
               EVp = 0*p1+50*p2+100*p3+150*p4
    )
    %>% mutate(DEV = EVq-EVp)
)
head(D)



```


Define and maximize likelihood:
```{r}

LL<-function(lambda,DEV,y) {
  # somthing in here to change DEV to DEU
  sum(-log(1+exp((-1)^y*lambda*DEV)))
}

LL(1,D$DEV,D$y)

lambdaHat<- optimize(function(x) {-LL(x,D$DEV,D$y)}, c(0, 1))

print(lambdaHat)

```

## Estimate a $\lambda_i$ for every subject (split by ID)
```{r}
lambdaHat<-function(DEV,y) {
  optimize(function(x) {-LL(x,DEV,y)}, c(0, 1))$minimum
}

print(lambdaHat(D$DEV,D$y))

LambdaHatEstimates<-(D
                     %>% group_by(id)
                     %>% summarize(lhat = lambdaHat(DEV,y)))


LambdaHatEstimates %>% head() %>% knitr::kable()

(
  ggplot(LambdaHatEstimates,aes(x=lhat))
  +stat_ecdf()
)
```

In context:
$$
\Pr(\text{chooses y to max DEV})=(1+\exp(-\lambda |\mathrm{DEV}|)^{-1}
$$
on average, a subject will choose the "right" action with probability
$$
\frac{1}{T}\sum_{t=1}^T(1+\exp(-\lambda |\mathrm{DEV}_t|)^{-1}
$$
```{r}
LogitResponse<-function(l,DEV) {
  mean(1/(1+exp(-l*abs(DEV))))
}
EVloss<-function(l,DEV) {
  p<-(1/(1+exp(-l*abs(DEV))))
  
  mean(p*0+(1-p)*abs(DEV))
}


LambdaHatEstimates<-(D
  %>% left_join(LambdaHatEstimates,by="id")
  %>% group_by(id)
  %>% summarize(PrHat = LogitResponse(lhat,DEV),
                EVloss = EVloss(lhat,DEV),
                lhat = mean(lhat))
)

LambdaHatEstimates %>% head() %>% knitr::kable()

(
  ggplot(LambdaHatEstimates,aes(x=PrHat))+stat_ecdf()
)

(
  ggplot()
  +stat_ecdf(data=LambdaHatEstimates,aes(x=EVloss,color="Estimated expected loss"))
  +stat_ecdf(data=D,aes(x=abs(DEV),color="Theoretical max"))
  +xlab("Expected loss relative to EV-maximizing (pounds)")
)

```

## Adding in risk aversion

Expected utility in vector form:
$$
\begin{aligned}
u(x)&=(u(x_1),u(x_2),\ldots)\\
p&=(p_1,p_2,p_3,p_4)\\
q&=(q_1,q_2,q_3,q_4)\\
E[u(X)\mid p]&=p^\top u(x)\\
E[u(X)\mid q]&=q^\top u(x)\\
DEU(u,p,q)&=(q-p)^\top u(x)
\end{aligned}
$$

```{r,cache=T}

prizes<-c(0,50,100,150)

LL<-function(params,p,q,y) {
  r<-exp(params[1])
  lambda<-exp(params[2])
  
  Ux<-as.matrix(prizes^r)
  
  
  EUq<-as.matrix(q) %*% Ux
  EUp<-as.matrix(p) %*% Ux
  
  
  DEU<-EUq-EUp
  
  
  sum(-log(1+exp((-1)^as.matrix(y)*lambda*DEU)))
}

p<-D %>%  dplyr::filter(id==1) %>% dplyr::select(p1:p4)
q<-D%>% dplyr::filter(id==1) %>% dplyr::select(q1:q4)
y<-D %>% dplyr::filter(id==1) %>% dplyr::select(y)


EUHat<-function(p,q,y) {
  exp(optim(c(log(0.5),log(0.7)),function(x) {-LL(x,p,q,y)})$par)
}

Estimates<-(D
                     %>% group_by(id)
                     %>% summarize(rhat = EUHat(p,q,y)[1],
                                   lhat = EUHat(p,q,y)[2])
            )


#Estimates %>% head() %>% knitr::kable()

library(latex2exp)
(
  ggplot(Estimates,aes(x=rhat,y=lhat))+geom_point()
  +theme_bw()
  #+xlab(TeX("$\\hat r$"))+ylab(TeX("$\\hat\\lambda$"))
   +ylim(c(0,exp(10)))
  +scale_y_continuous(trans="log10")
)


```

## Certainty equivalents: 

Lottery induced by behaving according to parameters $r$ and $\lambda$
$$
\begin{aligned}
\Pr(y_t=1\mid r,\lambda)&=(1+\exp(\lambda DEU_t(r)))^{-1}
\end{aligned}
$$
Induced lottery for choice $t$:
$$
q_t\Pr(y_t=1\mid r,\lambda)+p_t\left(1-\Pr(y_t=1\mid r,\lambda)\right)
$$
what is the lottery of the entire experiment?
$$
L=\frac{1}{T}\sum_{i=1}^T\left(q_t\Pr(y_t=1\mid r,\lambda)+p_t\left(1-\Pr(y_t=1\mid r,\lambda)\right)\right)
$$
Expected utility of experiment:
$$
EU(\text{expt})=L^\top u(x)
$$
Certainty equivalent of an experiment:
$$
CE(\text{expt})=u^{-1}\left(EU(\text{expt})\right)
$$

From here, we can calculate:

* How much is the experiment worth to someone who behaves according to $(r,\lambda)$?
* How much *could* it be worth if we shut down the noise (i.e. $\lambda\to\infty$)?
* How much was the experiment worth based on their actual choices?

## Hierarchical model

$$
\begin{aligned}
\log(r_i,\lambda_i)&\sim N(\mu,\Sigma)\\
\mu_r&\sim N(\log 0.5,0.2^2)\\
\mu_\lambda&\sim N(\log 10,0.2^2)\\
\log\sigma_r&\sim N(\log(0.2),0.2^2)\\
\log\sigma_\lambda&\sim N(\log(0.2),0.2^2)\\
\mathrm{corr}(\log r_i,\log\lambda_i)&\sim 2\mathrm{Beta}(3,3)-1
\end{aligned}
$$
```{r,eval=F}

d<-list(
  id=D$id,
  Q=D[,c("q1","q2","q3","q4")],
  P=D[,c("p1","p2","p3","p4")],
  y=D$y,
  N = dim(D)[1],
  n = length(unique(D$id)),
  prizes = seq(0,1,length=4),
  prior_r = c(log(0.5),0.2,log(0.2),0.2),
  prior_l = c(log(10),0.2,log(0.2),0.2),
  prior_corr = c(3,3)
  )

library(rstan)

Fit<-stan("EstimateCRRA.stan",data=d,control = list(adapt_delta=0.99))

saveRDS(Fit,"Hey2001Hierarchical.rds")

```


## Further reading

> Bland, J.R., 2020. Measuring and comparing two kinds of rationalizable opportunity cost in mixture models. Games, 11(1), p.1.

> Alekseev, Aleksandr, Glenn W. Harrison, Morten Lau, and Don Ross. "[Deciphering the noise: the welfare costs of noisy behavior](https://cear.gsu.edu/files/2018/03/WP_2018_01_Deciphering-the-Noise-The-Welfare-Costs-of-Noisy-Behavior_2019_0326.pdf)." (2019).