---
title: "Econometrics II Notes 20201"
author: "James Bland"
date: "Spring 2021"
output:
  bookdown::gitbook: default
---

```{r setup, include=FALSE}
library(ggplot2)
knitr::opts_chunk$set(echo = TRUE,warning = FALSE,message=FALSE,cache=TRUE)
```

# Limited dependent variable models

## Some example data

```{r loadTirtanic}
library(dplyr)
D<-(data.frame(read.csv("https://vincentarelbundock.github.io/Rdatasets/csv/carData/TitanicSurvival.csv"))
)

DSummary<-D %>% group_by(sex,age,passengerClass) %>% summarize(fracSurvived = mean(1*(survived=="yes")),n=n())

knitr::kable(head(DSummary))
```

```{r PlotTitanic,cache=T}
library(ggplot2)
(ggplot(DSummary ,aes(y=fracSurvived,x=age,color=sex,shape=passengerClass,size=n))
  +geom_point()
 +theme_bw()
  +xlab("Age")
  +ylab("Fraction survived")
  +theme_bw()
)
```

## Analyzing with linear regression

```{r linreg, results="asis"}

D <- (D
  %>% mutate(adult = 1*(age>=18),
             class2 = 1*(passengerClass=="2nd"),
             class3 = 1*(passengerClass=="3rd"),
             male = 1*(sex=="male"),
             S = 1*(survived=="yes")
             )
)

reg1<-lm(S ~ age + male+class2+class3,data=D)
reg2<-lm(S ~ age + adult+ male+class2+class3,data=D)
reg3<-lm(S ~ age + adult +adult*male + male + class2 + class3,data=D)


library(stargazer)
stargazer(reg1,reg2,reg3,type="html")

```

```{r PlotTitanicLinear,cache=T}

(ggplot(DSummary ,aes(y=fracSurvived,x=age,color=sex,shape=passengerClass,size=n))
  +geom_smooth(method="lm")
  +geom_point()
 +theme_bw()
  +xlab("Age")
  +ylab("Fraction survived")
  +theme_bw()
)
```

## Focusing on males only

```{r TitanicMale}

(
  ggplot(data=(D %>% filter(sex=="male")), aes(x=age,y=1*(survived=="yes"),group=paste(factor(age<18),passengerClass),color=passengerClass))
  +geom_smooth(method="lm",formula = y~ x + (1*(x<18)))
  +geom_point(data=(DSummary %>% filter(sex=="male")) ,aes(y=fracSurvived,x=age,color=passengerClass,size=n))
  
  +theme_bw()
  +labs(title="Titanic survival (males only)")
)

```

```{r,results="asis"}
Dmale<-(D%>%filter(sex=="male")
         %>% mutate(AgeMinus18 = age-18)
)

reg1<-lm(S ~ AgeMinus18 + class2+class3,data=Dmale)
reg2<-lm(S ~ AgeMinus18 + adult+class2+class3,data=Dmale)
reg3<-lm(S ~ AgeMinus18 + adult +AgeMinus18*adult+ class2 + class3,data=Dmale)

stargazer(reg1,reg2,reg3,type="html")
fitted<-(reg3$fitted.values)
(
  ggplot(data=data.frame(fitted),aes(x=fitted))
  +geom_histogram(bins=20)
  +theme_bw()
  +geom_vline(xintercept=0)
)

```

```{r,results="asis"}
probit1<-glm(S ~ AgeMinus18 + class2+class3,data=Dmale,family=binomial(link="probit"))
probit2<-glm(S ~ AgeMinus18+adult*(AgeMinus18) + adult+ class2+class3,data=Dmale,family=binomial(link="probit"))

stargazer(probit1,probit2,type="html")
```

## Marginal effects

Our goal is to evaluate:
$$
\frac{\partial \Pr(Y_i=1\mid X)}{\partial X_k} = 
\phi(\beta_0+\beta_1X_1+\beta_2X_2+\ldots)\beta_k
$$
### Average marginal effect
$$
\frac{1}{n}\sum_{i=1}^n\phi(\beta_0+\beta_1X_{1,i}+\beta_2X_{2,i}+\ldots)\beta_k
$$
```{r}
library(mfx)
Dmale<-Dmale %>% mutate(S = 1*(survived=="yes"))
p1<-glm(S ~ AgeMinus18,data=Dmale,family=binomial(link="probit"))
p1mfx<-probitmfx(formula=S~AgeMinus18, data=Dmale,atmean=FALSE)

print(p1mfx)

# replicate point estimate of these
B<-p1$coefficients

mfx<-dnorm(B[1]+Dmale$AgeMinus18*B[2])*B[2]
print(mean(mfx,na.rm=TRUE))
```

### Marginal effect at the mean
$$ \phi(\beta_0+\bar X_1\beta_1+\bar X_2\beta_2+\ldots)\beta_k  $$

```{r}
p1mfxatmean<-probitmfx(formula=S~AgeMinus18, data=Dmale,atmean=TRUE)

print(p1mfxatmean)

mfxatmean<-dnorm(B[1]+mean(Dmale$AgeMinus18,na.rm=TRUE)*B[2])*B[2]
print(mfxatmean)
```
### Marginal effect of an interaction

Continuous $X$, dummy $D$

$$
\begin{aligned}
\Pr(Y_i=1\mid X,D)&=\Phi(\beta_0+\beta_1X+\beta_2D+\beta_3XD)\\
\frac{\partial \Pr(Y_i=1\mid X,D)}{\partial X}&=\phi(\beta_0+\beta_1X+\beta_2D+\beta_3XD)(\beta_1+\beta_3D)\\
&=\begin{cases}
\phi(\beta_0+\beta_1X+\beta_2D+\beta_3XD)\beta_1 &\text{ if } D=0\\
\phi(\beta_0+\beta_1X+\beta_2D+\beta_3XD)(\beta_1+\beta_3) &\text{ if } D=1
\end{cases}
\end{aligned}
$$

```{r}

D<-D%>% mutate(S=1*(survived=="yes"),female=1*(sex=="female"),AgeMinus18=age-18)
p2<-glm(S ~ AgeMinus18+AgeMinus18*female,data=D,family=binomial(link="probit"))
print(p2)
# average marginal effect of age
B<-p2$coefficients

mfx1<-mean(dnorm(B[1]+B[2]*D$Ageminus18+B[3]*D$female+B[4]*D$Ageminus18*D$female)*(B[2]+D$female*B[4]),na.rm=TRUE)

print(mfx1)

# let's see what the mfx package gives us


mfxp2<-probitmfx(S ~ AgeMinus18+AgeMinus18*female,data=D,atmean=FALSE)
mfxp2a<-probitmfx(S ~ AgeMinus18+AgeMinus18*as.factor(female),data=D,atmean=FALSE)
print(mfxp2)
print(mfxp2a)

```

These are different, neither of the mfx package options work.

### Marginal effect for a dummy variable

$$
\begin{aligned}
\Pr(Y_i=1\mid X,D)&=\Phi(\beta_0+\beta_1X + \beta_2D)
\\
\Phi(\beta_0+\beta_1X + \beta_2)-\Phi(\beta_0+\beta_1X)
\end{aligned}
$$
```{r}
p3<-glm(S ~ AgeMinus18+female,data=D,family=binomial(link="probit"))
mfx3<-probitmfx(S ~ AgeMinus18+as.factor(female),data=D,atmean=FALSE)

print(mfx3)
```

## Maximizing a function

Coin flips

$$
\begin{aligned}
y_i&\sim iid \mathrm{Bernoulli}(\theta)\\
\mathcal L(\theta)&=n_1\log\theta+n_0\log(1-\theta)\\
\frac{\partial \mathcal L(\theta)}{\partial \theta}&=\frac{n_1}{\theta}-\frac{n_0}{1-\theta}\\
\frac{\partial^2\mathcal L(\theta)}{\partial \theta^2}&=-\frac{n_1}{\theta^2}-\frac{n_0}{(1-\theta)^2}
\end{aligned}
$$
Newton's method in general:
$$ 
\begin{aligned}
f(x)&\approx f(x_0)+(x-x_0)f'(x_0)+\frac12(x-x_0)^2f''(x_0)\\
\text{FOC: }0&=f'(x_0)+(x-x_0)f''(x_0)\\
x_{t+1}&=x_t-\frac{f'(x_t)}{f''(x_t)}
\end{aligned}
$$

In our example:
$$
\begin{aligned}
\theta_{t+1}&=\theta_t+\frac{\frac{n_1}{\theta_{t}}+\frac{n_0}{1-\theta_{t}}}{\frac{n_1}{\theta_t^2}+\frac{n_0}{(1-\theta_t^2)}}
\end{aligned}
$$

```{r, echo=T}

theta0<-0.1

n1<-1000
n0<-400

ThetaNext<-function(t0,n1,n0) {
  numerator<-n1/t0 - n0/(1-t0)
  denominator<-n1/t0^2+n0/(1-t0)^2
  f<-n1*log(t0)+n0*log(1-t0)
  
  ThetaNext<-t0+numerator/denominator
  
  c(ThetaNext,f,numerator,-denominator)
}

THETA<-seq(0.001,0.999,length=999)

D<-(data.frame(THETA)
    %>% mutate(Likelihood = n1*log(THETA)+n0*log(1-THETA))
)

thetaNewton<-theta0
derivatives<-c()

ApproxFun<-data.frame()

for (ii in 1:10) {
  f<-ThetaNext(theta0,n1,n0)
  
  x<-THETA[abs(THETA-theta0)<=0.2]
  
  y<-f[2]+(x-theta0)*f[3]+0.5*(x-theta0)^2*f[4]
  
  iteration<-ii-1
  
  theta0<-f[1]
  
  derivatives<-f[2:4]
  
  
  
  ApproxFun<-rbind(ApproxFun,data.frame(x,y,iteration))
  
  
  thetaNewton<-c(thetaNewton,theta0)
}

NewtonPoints<-data.frame(thetaNewton)%>% mutate(Likelihood = n1*log(thetaNewton)+n0*log(1-thetaNewton))
NewtonPoints$iteration<-0:(dim(NewtonPoints)[1]-1)



(ggplot(D,aes(x=THETA,y=Likelihood))
  +geom_path(size=1)
  +geom_point(data=NewtonPoints,aes(x=thetaNewton,y=Likelihood))
  +geom_path(data=ApproxFun,aes(x=x,y=y,group=iteration),color="red")
  +theme_bw()
  +ylim(c(-2000,0))
  
)

```

Algorithm does not converge

```{r}
n1<-1000
n0<-1

theta0<-0.1

thetaT<-theta0

for (ii in 1:10) {
  f<-ThetaNext(theta0,n1,n0)
  
  thetaT<-c(thetaT,f[1])
  theta0<-f[1]
}

print(thetaT)

```
Solution 1: don't move as far as the algorithms suggests:

```{r}
theta0<-0.1

n0<-10

thetaT<-theta0
c<-0.1

for (ii in 1:1000) {
  f<-ThetaNext(theta0,n1,n0)
  
  thetaNext<-f[1]*c +theta0*(1-c)
  theta0<-thetaNext
  thetaT<-c(thetaT ,thetaNext)
}

plot(thetaT)

```









