

# Stochastic dominance

## First-order stochastic dominance

$$
A \ \mathrm{FSD}\ B \iff p(A\leq x)\leq p(B \leq x)\quad \forall x
$$
One way to think about this is that lottery $A$ is lottery $B$ plus a bit more.

$$
A \sim B + Y, \quad \text{where the support of  } Y \text{ only includes nonnegative numbers}
$$

Example:
$$
A \sim N(0,1),\quad B\sim N(-1,1)
$$

If we conclude that $A$ FSD $B$, then anyone who likes more money to less money would prefer lottery $A$ to $B$

```{r}
library(tidyverse)
set.seed(42)

A<-rnorm(1000)
B<-rnorm(1000)-1
d<-data.frame(A,B)

(
        ggplot(data=d)
        +stat_ecdf(aes(x=A,color="A"))
        +stat_ecdf(aes(x=B,color="B"))
)

```

## Second-order stochastic dominance

$$
A \ \mathrm{SSD}\ B \iff \int_{-\infty}^xp(B\leq \tilde x)- p(A \leq \tilde x)\mathrm d\tilde x\geq 0\quad \forall x
$$
A special case of this is if:
$$
B = A + \epsilon-c,\quad E(\epsilon)=0,\quad p(\epsilon\mid A)=p(\epsilon), \quad c\geq 0 
$$

Can think of the special case as a "mean preserving spread", maybe minus a constant.

Example:
$$
A \sim N(0,1),\quad B\sim N(0,2)
$$

If we conclude that $A$ FSD $B$, then anyone who likes more money to less money would prefer lottery $A$ to $B$

```{r}
library(tidyverse)
set.seed(42)

A<-rnorm(1000)
B<-sqrt(2)*rnorm(1000)
d<-data.frame(A,B)

(
        ggplot(data=d)
        +stat_ecdf(aes(x=A,color="A"))
        +stat_ecdf(aes(x=B,color="B"))
)

```

If we conclude that $A$ SSD $B$, then anyone who is a risk-averse expected utility-maximizer will prefer $A$ to $B$

```{r}
library(dagitty)

g<-dagitty("
 dag {
 HI-> W
 Ability -> W
 Ability -> Job
 Ability->HI
 Job -> HI
 Husband->HI
 }          
"
)
plot(g)
print(paths(g,"HI","W"))
```

General conditions for an instrument to be valid:

* $\mathrm{cov}(Z,\epsilon)=0$
* $\mathrm{cov}(Z,X)\neq 0$