

# Predictions using structural models.

As an example, I will be using data from my paper:

 > Bland, James R. "[How many games are we playing? An experimental analysis of choice bracketing in games.](https://www.sciencedirect.com/science/article/pii/S2214804318303616?casa_token=-ItMAh0cDKsAAAAA:cgpnKmEY-aQis2gPnx0rz9FDbgUerQueArmhIDniBWh_fPqUknlt05jvbiQe1ufNBfU9SPwD6-Q)" Journal of Behavioral and Experimental Economics 80 (2019): 80-91.


## Lottery task 

**Part 1**

You have 100 tokens. You can invest some/all/none of it in a lottery. It will pay you $\chi$ times the amount you invested with probability 50%, zero otherwise.


$$
\text{payment}=\begin{cases}
100-y&\text{ if not successful}\\
100-y+\chi y&\text{otherwise}
\end{cases}
$$
**Part 2**

The same as Part 1, except that there were two independent lotteries.

**Goal:** Use Part 1 data to predict what will happen in Part 2 

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
library(kableExtra)
library(stringr)
D<- (read.csv("Data/HowManyGames.csv")
  %>% filter(Period==20)     
  %>% select(uid,contains("chi"),contains("invest"))
  %>% mutate(ID = as.integer(as.factor(paste(uid))))
)



chi1<-c(2.25,2.75,2.35,5.50)
chi2<-c(2.2,2.5,3.0,4.0)

Part1<-(D 
        %>% dplyr::select(ID,contains("invest1"))
        %>% gather(x,invest,invest1_1:invest1_4)
        %>% mutate(instance = as.integer(str_sub(x,-1,-1)),
                   Part = "Part 1")
        %>% select(ID,invest,instance,Part)
        %>% mutate(chi = chi1[instance])
)


Part2<-(D 
        %>% select(ID,contains("invest2"))
        %>% mutate(
          invest2_1=round(0.5*(invest2_1_1+invest2_2_1)),
          invest2_2=round(0.5*(invest2_1_2+invest2_2_2)),
          invest2_3=round(0.5*(invest2_1_3+invest2_2_3)),
          invest2_4=round(0.5*(invest2_1_4+invest2_2_4)),
          
          )
        %>% gather(x,invest,invest2_1:invest2_4)
        %>% mutate(instance = as.integer(str_sub(x,-1,-1)),
                   Part = "Part 2"
         )
        %>% select(ID,invest,instance,Part)
        %>% mutate(chi = chi2[instance])
        
)

LotteryData<-rbind(Part1,Part2)
print(LotteryData %>% sample_n(size=15))

```

```{r}


(
  ggplot(LotteryData,aes(x=invest,color=as.factor(instance)))
  +stat_ecdf()
  +facet_wrap(~Part)
  +theme_bw()
  +xlab("tokens invested")
  +ylab("ecdf")
  
)

```

## Model for Part 1:

$$
\begin{aligned}
u_i(x)&=x^{r_i}\\
p(Y_i=y)&\propto \exp(\lambda_i E[u(X)\mid y])\\
\log r_i&\sim N(\mu_r,\sigma_r^2)\\
\log\lambda_i&\sim N(\mu_\lambda,\sigma_\lambda^2)
\end{aligned}
$$
$$
\begin{aligned}
E[u(X)\mid y]&=0.5(100-y+\chi y)^r+0.5(100-y)^r\\
p(Y_i=y)&=\frac{\exp(\lambda_i E[u(X)\mid y])}{\sum_k\exp(\lambda_i E[u(X)\mid k])}
\end{aligned}
$$
Priors:

$$
\begin{aligned}
\mu_r&\sim N(\log(0.3),1)\\
\mu_\lambda&\sim N(\log(0.01),1)\\
\sigma_r&\sim N(\log(0.1),1)\\
\sigma_\lambda&\sim N(\log(0.1),1)\\
\end{aligned}
$$


```{r,eval=F,cache=F}
getwd()
library(rstan)
#options(mc.cores = parallel::detectCores())
options(mc.cores = 2)
rstan_options(auto_write = TRUE)


  #int<lower=0> N; // number of observations
  #int<lower=0> n; // number of participants
  #int invest[N];
  #vector[N] chi;
  #vector[N] Part;
  #vector[N] UseData;
  #int<lower=1> ID[N];
  #real priorR[4];
  #real priorL[4];
UseInvest<-Part1$invest+1
UseInvest[UseInvest==101]<-100


InvestGrid<-seq(0,100,length=11)
investL<-floor(Part1$invest/10)+1 
investR<-ceiling(Part1$invest/10)+1
investFrac<-(Part1$invest-InvestGrid[investL])/10
d<-list(
  N = dim(Part1)[1], 
  n = length(unique(Part1$ID)),
  investL = investL,
  investR = investR,
  investFrac=investFrac,
  nInvestGrid=length(InvestGrid),
  InvestGrid = InvestGrid,
  chi = Part1$chi,
  Part=1+1*(Part1$Part=="Part 1"),
  UseData=as.vector(matrix(1,dim(Part1)[1],1)),
  ID=Part1$ID,
  priorR=c(log(0.3),0.5,log(0.1),0.5),
  priorL=c(log(0.1),0.5)
)   
  
Fit<-stan("Bracketing.stan",data=d,control = list(adapt_delta = 0.999,max_treedepth=10))
saveRDS(Fit,"StructuralPredictions.rds")
```


```{r}
Fit<-readRDS("StructuralPredictions.rds")
print(Fit)
```
## Models for Part 2

**Narrow bracketing**: I ignore the first lottery when making my decision in the 2nd, and so on. The same decision problem as Part 1:

$$
\begin{aligned}
E[u(X)\mid y]&=0.5(100-y+\chi y)^r+0.5(100-y)^r
\end{aligned}
$$
**Broad bracketing**: I add up the prizes from both lotteries (what the "rational" model says you should do).

$$
\begin{aligned}
E[u(X_1+X_2)\mid y]&=0.25(200-2y+2\chi y)^r+0.5(200-2y+\chi y)^r+0.25(200-2y)^r
\end{aligned}
$$




